---
title: "Exploratory Analysis of Gonorrhea"
output: 
  html_document:
    code_folding: hide
---

```{r setup, include=FALSE}
library(tidyverse)
library(dplyr)
library(readxl)
library(readr)
library(plotly)
library(patchwork)

knitr::opts_chunk$set(
	fig.width = 11, 
  fig.asp = .6,
  out.width = "90%", 
	warning=FALSE
)
theme_set(theme_minimal() + theme(legend.position = "bottom"))
options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)
scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
```
  
```{r,  include = FALSE}
plots_df = read_csv("./data/clean_dataset.csv") %>%
  mutate(
    poverty_level = cut(poverty, breaks = c(-Inf, 10, 20, 30, 40, Inf), 
    labels = c("poverty_10","poverty_20", "poverty_30", "poverty_40", "poverty_40+"))) %>% 
  pivot_longer(
   cols = hispanic:other,
   names_to = "race",
   values_to = "percent_pop",
   values_drop_na = TRUE
 ) 
  
```
<br>

#### *The following plots are an exploration of the relationship between various predictors and the outcome of gonorrhea.*

<br>

### Part I: Health Insurance and Gonorrhea 

```{r}
gonorrhea1_ggplot = 
  plots_df %>% 
  ggplot(aes(x = health_ins, y = gonorrhea, group = nta_name)) + 
  geom_point(color = "green") + 
  labs(
    title = "Health Insurance and  Gonorrhea in NYC Neighborhoods (2014-2015)",
    x = "Percent population with health insurance",
    y = "Rate of gonorrhea cases per 100,000 people")

ggplotly(gonorrhea1_ggplot)
```

With the exception of a few outliers, the trend seems to be that neighborhoods with a lower percentage of population enrolled in health insurance have higher rates of gonorrhea cases. The association is not very strong but it is worth exploring further. 

<br>

### Part II: Medicaid and Gonorrhea 

```{r}
gonorrhea2_ggplot = 
  plots_df %>% 
  ggplot(aes(x = medicaid_enroll, y = gonorrhea, group = nta_name)) +
  geom_point(color = "green") + 
  labs(
    title = "Medicaid Enrollment and Gonorrhea in NYC Neighborhoods (2014-2015)",
    x = "Percent population enrolled in Medicaid",
    y = "Rate of gonorrhea cases per 100,000 people")

ggplotly(gonorrhea2_ggplot)

```

There is no strong relationship between percentage of population enrolled in Medicaid and rates of gonorrhea in the neighborhood population. However, there is a slight positive trend where neighborhoods with higher levels of Medicaid enrollment display higher rates of gonorrhea cases. There are a few outliers that do not follow this trend and indicate high rates of gonorrhea despite having low Medicaid enrollment. 

<br>

### Part III: Education and Gonorrhea

```{r}
gonorrhea3_ggplot = 
  plots_df %>% 
  ggplot(aes(x = education, y = gonorrhea, group = nta_name)) +
  geom_point(color = "green") + 
  labs(
    title = "Education Level and Gonorrhea in NYC Neighborhoods (2014-2015)",
    x = "Percent population with less than high school education",
    y = "Rate of gonorrhea cases per 100,000 people")


ggplotly(gonorrhea3_ggplot)
```

Similar to the association with Medicaid enrollment, the association between education level and rates of gonorrhea is not very strong. However, there is a slight positive trend where neighborhoods with higher percentages of the population holding less than high school level education also experience higher rates of gonorrhea.    

<br>

#### Education and Gonorrhea - Stratified by Race

```{r, include=FALSE}
race_df = read_csv("./data/clean_dataset.csv") %>% 
  mutate(
    hispanic_percent = cut(hispanic, breaks = c(-Inf, 20, 40, 60, 80, Inf), 
      labels = c("0-20%", "20-40%", "40-60%", "60-80%", "80-100%")),
    black_percent = cut(black, breaks = c(-Inf, 20, 40, 60, 80, Inf), 
      labels = c("0-20%", "20-40%", "40-60%", "60-80%", "80-100%")), 
    white_percent = cut(white, breaks = c(-Inf, 20, 40, 60, 80, Inf),
      labels = c("0-20%", "20-40%", "40-60%", "60-80%", "80-100%")),
    other_percent = cut(other, breaks = c(-Inf, 20, 40, 60, 80, Inf),
      labels = c("0-20%", "20-40%", "40-60%", "60-80%", "80-100%"))
    )
```


```{r, warning=FALSE}
smm_gonorrhea_hispanic = 
  race_df %>% 
  ggplot(aes(x = education, y = gonorrhea)) +
  geom_point(aes(color = hispanic_percent)) +
  labs(
    title = "Gonorrhea vs education by % Hispanic population",
    x = "% with less than high school education",
    y = "Gonorrhea",
    color = "Population:" 
  )

smm_gonorrhea_black = 
  race_df %>% 
  ggplot(aes(x = education, y = gonorrhea)) +
  geom_point(aes(color = black_percent)) +
  labs(
    title = "Gonorrhea vs education by % Black population",
    x = "% with less than high school education",
    y = "Gonorrhea",
    color = "Population:" 
  )

smm_gonorrhea_white = 
  race_df %>% 
  ggplot(aes(x = education, y = gonorrhea)) +
  geom_point(aes(color = white_percent)) +
  labs(
    title = "Gonorrhea vs education by % White population",
    x = "% with less than high school education",
    y = "Gonorrhea",
    color = "Population:" 
  )

smm_gonorrhea_other = 
  race_df %>% 
  ggplot(aes(x = education, y = gonorrhea)) +
  geom_point(aes(color = other_percent)) +
  labs(
    title = "Gonorrhea vs education by % Other race population",
    x = "% with less than high school education",
    y = "Gonorrhea",
    color = "Population:" 
  )

gonorrhea_education_all = smm_gonorrhea_hispanic + smm_gonorrhea_black + smm_gonorrhea_white + smm_gonorrhea_other +
  labs(caption = "*Rate of Gonorrhea per 100,000 people") +
  theme(plot.caption = element_text(hjust = 0, face = "italic"))

gonorrhea_education_all
```

Among populations with higher percentages of Black or Hispanic people, there appears to be higher levels of both those with less than high school education and gonorrhea. In populations with higher percentages of White people, the opposite appears to be true where there are lower rates of less than high school education and gonorrhea. 