---
title: "Preterm Births Regression"
output: 
  html_document:
    code_folding: hide
---

```{r setup, include=FALSE}
library(tidyverse)
library(dplyr)
library(readxl)
library(readr)
library(ggpubr)
library(corrplot)
library(RColorBrewer)
library(Hmisc)
library(patchwork)
library(rstatix)
library(glmnet)
library(MASS)

knitr::opts_chunk$set(
	fig.width = 9, 
  fig.asp = .6,
  out.width = "90%",
	warning = FALSE,
	message = FALSE
)
theme_set(theme_minimal() + theme(legend.position = "bottom"))
options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)
scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
```

Let's start by fitting a linear model to the preterm births outcome using a stepwise selection process (with AIC as metric). Next, we'll take a look at the chosen predictors, effect measures, and run regression diagnostics of our converged model.

```{r}
#read in cleaned csv files
predictors_df = read.csv("./data/predictors.csv")
outcome_df = read.csv("./data/outcomes.csv")

#Combined predictors with preterm births outcome
preterm_linear_df = predictors_df %>% 
  mutate(preterm_births = outcome_df$preterm_births)

#Fit the full model 
full_preterm_linear.model <- lm(preterm_births~., data = preterm_linear_df)
#Stepwise regression model
step_preterm_linear.model <- stepAIC(full_preterm_linear.model, direction = "both", 
                      trace = FALSE)


#Display converged model
step_preterm_linear.model %>% 
  broom::tidy() %>%
  knitr::kable(digits = 3)

```
The 6 variables hispanic, black, limited_eng, health_ins, late_no_prenatal_care, and non_pcmd_density were chosen in the final model. The model has an R-squared value of `summary(step_preterm_linear.model)$r.squared`, representing a decent, but nowhere near ideal fit with our predictor set

However, we were concerned about our assumptions for linear regression, so they need to be checked before we can interpret any results. 
```{r}
#Regression diagnostics
plot(step_preterm_linear.model)

#Shapiro test
preterm_linear_shapiro = shapiro.test(residuals(step_preterm_linear.model))
```
The linear regression assumptions are satisfied, so this is a valid model. The Normal Q-Q plot is reasonably well fit, and Shapiro-Wilk test for normality, which has a p-value of `preterm_linear_shapiro$p.value`, confirms the normality assumption. The data also roughly follows the homoskedasticity assumption, with a few potential outliers right at 3 standard deviations and potential influential points. Though several point has high leverage, it is not higher than our rule-of-thumb threshold of 0.074. This is calculated using the rule of thumb equation 2(p + 1)/n, where p is the number of predictors and n the number of observations.

These results are valid, so we can interpret them!


