---
title: "Project Report"
output: 
  html_document:
    code_folding: hide
---

```{r setup, include=FALSE}
library(tidyverse)
library(dplyr)
library(readxl)
library(readr)
```
## Motivation 
Recent events in American politics - a surging pro-life movement, an increasingly conservative Supreme Court, and pending litigation against the Affordable Care Act (ACA) will potentially restrict access to reproductive services & resources for women across the country.

We seek to highlight the importance of such services by demonstrating how increased access to reproductive services lowers rates of STDs/STIs and maternal mortality in New York City, and how these health outcomes vary by socioeconomic factors like income and race.


## Questions
- How does access to reproductive services impact maternal mortality rates and levels of STIs? 
- Insurance
- How do socioeconomic factors of health like income and race affect the relationship between access to services and health outcomes? 


## Epidemiology 
The following are our directed acyclic graphs (DAGs) to visually represent the causal assumptions of our project and help identify the presence of confounding for our questions.


## Data
*Source:* 
[NYC Neighborhood Health Atlas](https://www1.nyc.gov/site/doh/health/neighborhood-health/nyc-neighborhood-health-atlas.page)

*Variables of interest*
Outcomes:

- `preterm births`: percent preterm births (less than 37 weeks gestation) among all live births
- `SMM`: crude rate of severe maternal morbidity (SMM) per 10,000 deliveries
- `gonorrhea`: crude rate of gonorrhea cases diagnosed in 2014-2015 per 100,000 population, all ages


Predictors:

- `health insurance coverage`: 
- `access to health services `: 
- `race`: 
- `poverty`: 
- `education `: 

## Data Tidying

The following steps were taken to clean the NYC NTA data: 

  * Clean names to improve readability
  * Select relevant predictors & outcomes (demographics, SES, health outcomes, insurance status, number of providers)
  * Drop neighborhoods that have no residential population (i.e. LaGuardia airport)
  * Drop observation representing NYC aggregated
  * Remove variables corresponding to 95% CI and reliability metrics of measurements
  * Create new variable clinics_and_chc_count representing aggregated clinic and community health centers to reduce multicollinearity
  * Create new variables adjusting number of healthcare providers by neighborhood population
      + clinics_and_chc_density: clinic & community health centers per 10,000 people
      + non_pcmd_density:  on-primary care providers (specialists like ob-gyns) per 10,000 people
  * Saved cleaned dataset to CSV to improve portability across users & programs
  
```{r}
project_df = read_excel("./data/nta-metadata.xlsx", sheet = "NTA Data") %>%
  janitor::clean_names() %>%
  select(nta_name:female, hispanic:limited_eng,edu_less_than_hs:poverty,unemployment,preterm_births:smm_reliability, gonorrhea, health_ins, medicaid_enroll, late_no_prenatal_care, clinics, community_health_centers, non_pcmd, -contains("95cl"), -contains("reliability")) %>%
  mutate(clinics_and_chc_count = clinics + community_health_centers,
         clinics_and_chc_density = ((clinics + community_health_centers)/total_pop)*10000,
         non_pcmd_density = (non_pcmd/total_pop)*10000) %>%
  drop_na(total_pop) %>%
  filter(nta_code != 'NYC')

write.csv(project_df, 'data/clean_dataset.csv', row.names = FALSE)
```


