---
title: "Project Report"
output: 
  html_document:
    code_folding: hide
---

```{r setup, include=FALSE}
library(tidyverse)
library(dplyr)
library(readxl)
library(readr)
```

## Motivation 
Recent events in American politics - a surging pro-life movement, an increasingly conservative Supreme Court, and pending litigation against the Affordable Care Act (ACA) will potentially restrict access to reproductive services & resources for women across the country.

We seek to highlight the importance of such services by demonstrating how increased access to reproductive services lowers rates of STDs/STIs and maternal morbidity in New York City, and how these health outcomes vary by socioeconomic factors like income and race.

<br>

## Questions

Among neighborhoods in NYC...

### Exploratory Analysis
- How does access to reproductive services impact maternal health and levels of STIs? 
- How does insurance coverage affect maternal health and levels of STIs?
- How is the relationship between maternal & reproductive outcomes and access to healthcare affected by factors like neighborhood racial composition, poverty and education?

### Neighborhood-Level Maps
- Spatial distributions of key demographic and socioeconomic predictors of maternal and reproductive among NYC neighborhoods
- Neighborhood clustering patterns to identify areas of particularly high or low risk

### Regression Models
- How do maternal and reproductive health outcomes correlate with demographic, socioeconomic, and healthcare access factors?
- Can we build accurate estimation models to quantify the relationship between health outcomes and their potential protective & risk factors?

<br>

## Data
#### **Source:** \ [NYC Neighborhood Health Atlas](https://www1.nyc.gov/site/doh/health/neighborhood-health/nyc-neighborhood-health-atlas.page)

&nbsp;

### *Variables of interest*
#### **Outcomes:**

- `preterm_births`: percent preterm births (<37 weeks gestation) among all live births in 2010-2014
- `smm`: crude rate of severe maternal morbidity (SMM) per 10,000 deliveries in 2008-2012
- `gonorrhea`: crude rate of gonorrhea cases diagnosed per 100,000 people of all ages in 2014-2015 

#### **Predictors:**

- `health_ins`: percent of civilian noninstitutionalized population with health insurance in 2010-2014
- `medicaid_enroll`: percent of population continuously enrolled (for 11 months or more) in Medicaid in 2015
- `late_no_prenatal_care`: percent of live births receiving late prenatal care or no prenatal care at all in 2010-2014
- `race`: various race variables were re-categorized into single variable with four categories: black, hispanic, white, and other
- `poverty`: percent of population all ages living below the federal poverty level. This variable was re-categorized into five levels during our exploratory analyses: 
  + `poverty_10`: less than 10% of population living below the federal poverty level
  + `poverty_20`: between 10-20% of population living below the federal poverty level
  + `poverty_30`: between 20-30% of population living below the federal poverty level
  + `poverty_40`: between 30-40% of population living below the federal poverty level
  + `poverty_40+`: greater than 40% of population living below the federal poverty level
- `education`: percent of population ages 25 and older whose highest level of education is less than a high school diploma or GED

<br>

## Data Tidying

The following steps were taken to clean the NYC NTA data: 

  * Clean names to improve readability
  * Select relevant predictors & outcomes (demographics, SES, health outcomes, insurance status, number of providers)
  * Drop neighborhoods that have no residential population (i.e. LaGuardia airport)
  * Drop observation representing NYC aggregated
  * Remove variables corresponding to 95% CI and reliability metrics of measurements
  * Create new variable clinics_and_chc_count representing aggregated clinic and community health centers to reduce multicollinearity
  * Create new variables adjusting number of healthcare providers by neighborhood population
      + `clinics_and_chc_density`: clinic & community health centers per 10,000 people
      + `non_pcmd_density`: on-primary care providers (specialists like ob-gyns) per 10,000 people
  * Renamed variables for ease of reading
  * Combined Asians into the 'other' race category
  * Saved cleaned dataset to CSV to improve portability across users & programs

```{r}
project_df = read_excel("./data/nta-metadata.xlsx", sheet = "NTA Data") %>%
  janitor::clean_names() %>%
  dplyr::select(
    nta_name:female, 
    hispanic:limited_eng,
    edu_less_than_hs:poverty,
    unemployment,
    preterm_births:smm_reliability, 
    gonorrhea, 
    health_ins, 
    medicaid_enroll, 
    late_no_prenatal_care, 
    clinics, 
    community_health_centers, 
    non_pcmd, 
    -contains("95cl"), 
    -contains("reliability")) %>%
  mutate(clinics_and_chc_count = clinics + community_health_centers,
         clinics_and_chc_density = ((clinics + community_health_centers)/total_pop)*10000,
         non_pcmd_density = (non_pcmd/total_pop)*10000,
         other = (asian_pi + other_race)) %>%
  rename(
    white = white_non_hisp,
    black = black_non_hisp,
    education = edu_less_than_hs) %>% 
  relocate(other, .after = black) %>% 
  select(-asian_pi, -other_race) %>% 
  drop_na(total_pop) %>%
  filter(nta_code != 'NYC')

write.csv(project_df, 'data/clean_dataset.csv', row.names = FALSE)
```
Our final dataset contained data from 188 neighborhoods in New York City, defined by Neighborhood Tabulation Areas (NTAs), which were generated from 2010 US Census data

## Methods

In this section, we discuss the analyses we completed and our results

### Exploratory Analysis


### Neighborhood-Level Maps

### Regression Models

#### Outcome

#### Outcome

#### Outcome

## Results & Discussion

## Acknowledgements

Special thanks to:

* Professor Qixuan Chen & Dr. Dan Liao for their statistics expertise
* Dr. Jeremy Porter for his advice on geospatial information systems (GIS)
* Jeff and the P8105 course staff (with special thanks to our TA, Junxian Chen) for showing us the endless possibilities of R
* The extraordinary, powerful, nasty women who inspire us every day






