---
title: "Project Report"
output: 
  html_document:
    code_folding: hide
---

```{r setup, include=FALSE}
library(tidyverse)
library(dplyr)
library(readxl)
library(readr)
```

## Introduction
Severe maternal morbidity (SMM) includes health outcomes during labor and delivery that may result in both short-term and/or long term health consequences for women. New York City’s rate of SMM increased 28.2 % from 2008 to 2012. Some of the factors contributing to high rates of SMM in New York City may include access to reproductive health services, rates of sexually transmitted infections (STIs), neighborhood poverty level and race. We will examine the factors that may contribute to the high SMM rates in New York City. To further assess the status of maternal health in nyc, we will examine the outcome of preterm births which are babies born less than 37 weeks of gestation. In regard to STI's in our assessment of reproductive health in NYC, we will choose to focus specifically on gonorrhea as it is provides considerable risk to pregnant mothers and newborn babies as well [1](https://www.mayoclinic.org/diseases-conditions/gonorrhea/symptoms-causes/syc-20351774#:~:text=Gonorrhea%20is%20an%20infection%20caused,vaginal%2C%20oral%20or%20anal%20sex.).  

<br>

## Literature Review

**Severe Maternal Morbidity** 

Maternal morbidity is defined by the WHO as "any health condition attributed to and/or aggravated by pregnancy and childbirth that has a negative impact on the woman’s well-being”. The medical causes of morbidity include hemorrhaging and pregnancy related infections, induced abortions, and complications during labor. Outside of these medical factors, there are many socioeconomic determinants to rates of maternal morbidity within a population. According to literature these most often include access to quality maternal health services, both during pregnancy and during labor. Access may come in the form of health insurance, proximity to reproductive health care facilities, and quality of care available at health care facilities [1](https://www.ncbi.nlm.nih.gov/books/NBK361917/). In New York City, racial and ethnic disparities have contributed significantly to high rates of maternal morbidity. According to a study published in the Maternal and Child Health Journal, SMM rates were high among black, non-latina women in particular. Maternal education, income, and lack of prenatal care were also significantly associated with SMM [2](https://link-springer-com.ezproxy.cul.columbia.edu/article/10.1007/s10995-018-2682-z).  

**Preterm Births**

Preterm birth is defined as the live birth of an infant prior to 37 weeks of gestation, with the categories of late preterm at 34-36 weeks, moderately preterm at 32-36 weeks, and very preterm at less than 32 weeks [1](https://www.marchofdimes.org/peristats/ViewTopic.aspx?reg=36&top=3&lev=0&slev=4). In addition to putting babies at risk of underdevelopment of vital organs, preterm births also affect maternal health by increasing the mother’s risk of developing postnatal physical health problems, depression, and lower usage of postnatal support [2](https://bmjopen.bmj.com/content/6/10/e012676), [3](https://www.frontiersin.org/articles/10.3389/fpubh.2018.00359/full). Racial and ethnic disparities exist in preterm birth rates across the nation, with the highest rates of preterm births among African-American mothers (14.4%) that are almost 50% higher than the rates among white (9.3%) or Hispanic (10%) women in 2019 [4](https://www.cdc.gov/reproductivehealth/maternalinfanthealth/pretermbirth.htm). In NYC, the overall rate of preterm births was 9.0% in 2017. When categorized by race, the preterm birth rate was 12.7% among Black women, 9.6% among Hispanic women, 7.9% among Asian women, 6.9% among White women, and 9.9% among women of Other race [5](https://data.cccnewyork.org/data/table/1276/preterm-births#1277/1480/25/a/a). While there are multiple risk factors for preterm birth including previous preterm births, pregnancy with multiples, tobacco and substance usage, short durations between pregnancies, and pregnancy complications, the socioeconomic environment of the mother is also important [6](https://www.cdc.gov/reproductivehealth/features/premature-birth/). In a 2008 International Journal of Epidemiology report, maternal educational level was found to be the strongest predictor for preterm birth out of five socioeconomic factors, more so than income and occupation [7](https://watermark.silverchair.com/dyn112.pdf?token=AQECAHi208BE49Ooan9kkhW_Ercy7Dm3ZL_9Cf3qfKAc485ysgAAApwwggKYBgkqhkiG9w0BBwagggKJMIIChQIBADCCAn4GCSqGSIb3DQEHATAeBglghkgBZQMEAS4wEQQMTy4eP8rA23acKExiAgEQgIICT7nfOcXpJt31qdbBKOIboYVrwnd_WHQhBeR2zPtf2_4u3bePmf8VV6qV0fEXrHgXvb3N2XB4sesbpMG_2si0irr_CROVEu1ViSWoRkPfTWYpg4tzoXjDr6sR8yZBM-Ka_4WEOclHYN3-guKHuT5Xtrm0cLFG4dcV8cv8xodawfaFfug75W9choeMXMKMea9ndCCyFL0xUuMLS16cXGVHSYzSixgj2MK1vDWLB-jt973iC34bxIBwwTP39ibenmDoH5qD8g6KphGfp8V5ANpwJofRkzYf3CKu9PbMeQzEqOOmTfShwfSn0Wi7jeR8Q37ZwznMKHiFrChO5ndZ46cTAY6az5Dw-ryrQP-S7yOK8t_M9a90Tcj6sUmhsie3WusXcASh_tlI2NySIF7sqMGjWuo_tybivPEpyxjpZUaNd-zaXLmI_wXbhtFWv4UsxzOVi8xG82P3BfvqImmsdkKJT2K28fC6qdBU5ee1Jo7jwXAEtAFpa262U5O1x64gQr76VZ15Z3QSQvYnG6YuJBjq1U3PGYN6svysasziSMNghnwNFgKZroCkWoyv2xe_FBKBP5Ju9Wt5ivOzkT9TLKlfw1n70UmqImeO4CQpdIYafasooL4ijLl_el-A7f5RvP_T9hvdAJ-7Sm-iI_c93luRWtsFJ46JDh4fkADgPRRp41ktocLFPP1KnHxE_diJbS-GscLBxE6jlxRsk8VdObQldz_VSoHw4lRQS0R05dbxcvlVtURdqJeAj8hXgtJLvmCPt9zaErgQZTaDO9bKlxJ8yQ). System-level factors are also seen to affect preterm birth, including the amount of prenatal care, management of maternal chronic disease, and health insurance coverage [8](https://www.tandfonline.com/doi/full/10.1080/10796126.2016.1254601). 

**Gonorrhea**

STI transmission and prevention is a big part of reproductive health. Gonorrhea in particular is a sexually transmitted infection that affects both men and women. It causes infections in the genitals, rectum, and throat and is treatable with antibiotics. While it is a sexually transmitted disease, gonorrhea can also be spread from mother to child during childbirth [1](https://www.cdc.gov/std/gonorrhea/stdfact-gonorrhea.htm). New York City has one of the highest rates of gonorrhea infections in the United States and this rate has almost doubled between 2010 and 2017. In 2017, there were 23,491 cases reported in NYC which was a 23.4% increase from 2016. In part, this is due to an increased use of pre-exposure prophylaxis which has led to decreased condom use. This behavior change has allowed other STI's, like gonorrhea, to increase [2](https://www.politico.com/states/new-york/albany/story/2018/09/25/in-nyc-gonorrhea-chlamydia-cases-up-syphilis-down-625004#:~:text=There%20were%2023%2C491%20cases%20of,percent%20from%20the%20prior%20year.). Despite these increasing trends, most of the new cases are amongst men, not women [3](https://www.health.ny.gov/statistics/diseases/communicable/std/docs/sti_surveillance_report_2017.pdf). At a national level, rates of gonorrhea remain highest among the black population (7.7 times the rate amongst the white population in 2018) [4](https://www.cdc.gov/std/stats18/STDSurveillance2018-full-report.pdf). Overall, there is not a great body of literature that shows population level relationships, instead trends are focused at the individual level.  

<br>

## Inspiration



## Questions

Among neighborhoods in NYC...

### Exploratory Analysis
- How does access to reproductive services impact maternal health and levels of STIs? 
- How does insurance coverage affect maternal health and levels of STIs?
- How is the relationship between maternal & reproductive outcomes and access to healthcare affected by factors like neighborhood racial composition, poverty and education?

### Neighborhood-Level Maps
- Spatial distributions of key demographic and socioeconomic predictors of maternal and reproductive among NYC neighborhoods
- Neighborhood clustering patterns to identify areas of particularly high or low risk

### Regression Models
- How do maternal and reproductive health outcomes correlate with demographic, socioeconomic, and healthcare access factors?
- Can we build accurate estimation models to quantify the relationship between health outcomes and their potential protective & risk factors?

<br>

## Data
#### **Source:** \ [NYC Neighborhood Health Atlas](https://www1.nyc.gov/site/doh/health/neighborhood-health/nyc-neighborhood-health-atlas.page)

<br>

### *Variables of interest*
#### **Outcomes:**

- `preterm_births`: percent preterm births (<37 weeks gestation) among all live births in 2010-2014
- `smm`: crude rate of severe maternal morbidity (SMM) per 10,000 deliveries in 2008-2012
- `gonorrhea`: crude rate of gonorrhea cases diagnosed per 100,000 people of all ages in 2014-2015 

#### **Potential Predictors:**

- `nta_name`: 
- `nta_code`: 
- `total_pop`:
- `female`: percent of female population in 2010-2014
- `foreign_born`: percent of population born outside the US or US territories
- `limted_eng`:
- `unemployment`:
- `health_ins`: percent of civilian noninstitutionalized population with health insurance in 2010-2014
- `medicaid_enroll`: percent of population continuously enrolled (for 11 months or more) in Medicaid in 2015
- `late_no_prenatal_care`: percent of live births receiving late prenatal care or no prenatal care at all in 2010-2014
- `race`: various race variables were re-categorized into a single variable with four categories: black, hispanic, white, and other
- `poverty`: percent of population all ages living below the federal poverty level. This variable was re-categorized into five levels during our exploratory analyses: 
  + `poverty_10`: less than 10% of population living below the federal poverty level
  + `poverty_20`: between 10-20% of population living below the federal poverty level
  + `poverty_30`: between 20-30% of population living below the federal poverty level
  + `poverty_40`: between 30-40% of population living below the federal poverty level
  + `poverty_40+`: greater than 40% of population living below the federal poverty level
- `education`: percent of population ages 25 and older whose highest level of education is less than a high school diploma or GED
- `clinics`:
- `community_health_centers`:
- `non_pcmd`: 
- `clinics_and_chc_count`:
- `clinics_and_chc_density`:
- `non-pcmd_density`:

<br>

## Data Tidying

The following steps were taken to clean the NYC NTA data: 

  * Clean names to improve readability
  * Select relevant predictors & outcomes (demographics, SES, health outcomes, insurance status, number of providers)
  * Drop neighborhoods that have no residential population (i.e. LaGuardia airport)
  * Drop observation representing NYC aggregated
  * Remove variables corresponding to 95% CI and reliability metrics of measurements
  * Create new variable clinics_and_chc_count representing aggregated clinic and community health centers to reduce multicollinearity
  * Create new variables adjusting number of healthcare providers by neighborhood population
      + `clinics_and_chc_density`: clinic & community health centers per 10,000 people
      + `non_pcmd_density`: on-primary care providers (specialists like ob-gyns) per 10,000 people
  * Renamed variables for ease of reading
  * Combined Asians into the 'other' race category
  * Saved cleaned dataset to CSV to improve portability across users & programs

```{r}
project_df = read_excel("./data/nta-metadata.xlsx", sheet = "NTA Data") %>%
  janitor::clean_names() %>%
  dplyr::select(
    nta_name:female, 
    hispanic:limited_eng,
    edu_less_than_hs:poverty,
    unemployment,
    preterm_births:smm_reliability, 
    gonorrhea, 
    health_ins, 
    medicaid_enroll, 
    late_no_prenatal_care, 
    clinics, 
    community_health_centers, 
    non_pcmd, 
    -contains("95cl"), 
    -contains("reliability")) %>%
  mutate(clinics_and_chc_count = clinics + community_health_centers,
         clinics_and_chc_density = ((clinics + community_health_centers)/total_pop)*10000,
         non_pcmd_density = (non_pcmd/total_pop)*10000,
         other = (asian_pi + other_race)) %>%
  rename(
    white = white_non_hisp,
    black = black_non_hisp,
    education = edu_less_than_hs) %>% 
  relocate(other, .after = black) %>% 
  select(-asian_pi, -other_race) %>% 
  drop_na(total_pop) %>%
  filter(nta_code != 'NYC')

write.csv(project_df, 'data/clean_dataset.csv', row.names = FALSE)
```
Our final dataset contained data from 188 neighborhoods in New York City, defined by Neighborhood Tabulation Areas (NTAs), which were generated from 2010 US Census data

## Methods

In this section, we discuss the analyses we completed and our results

### Exploratory Analysis

We used data visualizations to conduct some exploratory analysis of our outcomes and predictors to identify any preliminary trends. We approched this by separating the three primary outcomes (SMM, preterm births, and gonorrhea) and selected various predictors based on literature to further examine their relationships. In order to create the plots, we first took our clean data set and restructured the poverty and race predictors into categorical variables. Then, for each outcome we created scatter plots against predictors such as race, poverty, prenatal care, health insurance, medicaid enrollment, and education. For gonorrhea, we also investigated its relationship with education, stratified by race. Similarly, we investigated SMM and prenatal care, stratified by race. 

### Neighborhood-Level Maps

### Regression Models

#### Outcome

#### Outcome

#### Outcome

## Results & Discussion

## Acknowledgements

Special thanks to:

* Professor Qixuan Chen & Dr. Dan Liao for their statistics expertise
* Dr. Jeremy Porter for his advice on geospatial information systems (GIS)
* Jeff and the P8105 course staff (with special thanks to our TA, Junxian Chen) for showing us the endless possibilities of R
* The extraordinary, powerful, nasty women who inspire us






