---
title: "Regression Overview"
output: 
  html_document:
    code_folding: hide
---

```{r setup, include=FALSE}
library(tidyverse)
library(dplyr)
library(readxl)
library(readr)
library(ggpubr)
library(corrplot)
library(RColorBrewer)
library(Hmisc)
library(patchwork)
library(rstatix)
library(glmnet)
library(MASS)

knitr::opts_chunk$set(
	fig.width = 9, 
  fig.asp = .6,
  out.width = "90%",
	warning = FALSE,
	message = FALSE
)
theme_set(theme_minimal() + theme(legend.position = "bottom"))
options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)
scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
```
## Summary
[summary of results goes here]

## Data

#### **Source:** \ [NYC Neighborhood Health Atlas](https://www1.nyc.gov/site/doh/health/neighborhood-health/nyc-neighborhood-health-atlas.page)

The NTA Neighborhood data contains data from 188 neighborhoods in New York City, defined by Neighborhood Tabulation Areas (NTAs), which were generated from whole census tracts from 2010 US Census data. Created by aggregating adjacent census tracts, the NTA neighborhood dataset is a happy medium between the high sampling error of NYC's overly granular 2,168 census tracts and the broad strokes of the 59 community districts. However, because their primary boundary creation consideration is population, NTAs do not accurately represent the boundaries of NYC's historic neighborhoods.

&nbsp;

### *Variables of interest*
#### **Outcomes:**

- `preterm_births`: percent preterm births (<37 weeks gestation) among all live births in 2010-2014
- `smm`: crude rate of severe maternal morbidity (SMM) per 10,000 deliveries in 2008-2012
- `gonorrhea`: crude rate of gonorrhea cases diagnosed per 100,000 people of all ages in 2014-2015 

#### **Predictors:**

15 potential predictor variables were identified after cleaning and transformation. Predictor selection was driven by subject matter knowledge, literature review, dataset constraints, and common sense.

## Analysis Plan

We will begin with assuming a linear relationship between predictors and outcome and move to more complex relationships via transformation if a linear model is a poor fit. 

Models will be selected with a stepwise automatic model selection process, which iteratively adds and removes predictors to the model until it converges, using AIC as a metric. Please note that there is some debate regarding the statistical validity of stepwise model selection, further discussed in the appendix. For our purposes, we will use it for exploratory analysis due to ease of implementation and its intuitive approach and approach our results critically with respect to the algorithm's pitfalls.

Models will be judged on:
  * Satisfaction of assumptions for their respective method
  * Best fit as judged by Akaike information criterion (AIC), which considers both goodness-of-fit (rewarding models that explain higher variation in the outcome) and model parsimony (penalizing for too many predictors). AIC is applicable to many distributions, including linear, Poisson, and negative binomial.
  
### Outcome Variables

We have three outcome variables of interest: preterm births, severe maternal morbidity (SMM), and gonorrhea. 

We start with univariate analyses dependent variables using histograms, Q-Q plots and Shapiro tests for normality. Ultimately, it is the residual distribution that matters in multivariable regression, but it never hurts to visualize data.

```{r}
#read in data
project_df = read.csv("./data/clean_dataset.csv")
#separate outcomes into separate dataframe
outcome_df = project_df %>%
  dplyr::select(preterm_births, smm, gonorrhea)

write.csv(outcome_df, 'data/outcomes.csv', row.names = FALSE)

#histograms
preterm_hist = ggplot(outcome_df, aes(x = preterm_births)) + geom_histogram()
smm_hist = ggplot(outcome_df, aes(x = smm)) + geom_histogram()
gonorrhea_hist = ggplot(outcome_df, aes(x = gonorrhea)) + geom_histogram()

#q-q plots
preterm_qq = ggqqplot(outcome_df$preterm_births)
smm_qq = ggqqplot(outcome_df$smm)
gonorrhea_qq = ggqqplot(project_df$gonorrhea)

#create patchwork plots
preterm_hist + preterm_qq

smm_hist + smm_qq

gonorrhea_hist + gonorrhea_qq

#run shapiro-wilk tests
outcome_df %>% shapiro_test(preterm_births, smm, gonorrhea) %>%
  knitr::kable(digits = 3)
```

Outcome variables are likely sampled from a non-normal distribution, as indicated by significant p-values for the Shapiro test and fat tails in the Q-Q plots: not a deal-breaker, since residuals are what are important, but certainly presents the possibility that linear regression may not be the best model for our data.

### Multicollinearity Considerations

Before we start building models, let's check for potential multicollinearity among predictors, since stepwise model selection will exacerbate any existing multicollinearity in our model. 

To summarize simple correlations between variables, we'll construct Pearson correlation matrix of predictors, using pair-wise (instead of case) deletion of datapoints with missing values.

```{r}
#create dataset of predictors ONLY
predictors_df = project_df %>% 
  dplyr::select(-c("nta_name","nta_code", "preterm_births", "gonorrhea", "smm", "clinics", "community_health_centers", "non_pcmd", "clinics_and_chc_count"))

write.csv(predictors_df, 'data/predictors.csv', row.names = FALSE)

N <- rcorr(as.matrix(predictors_df), type = c("pearson")) #pair-wise deletion

corrplot::corrplot(N$r,
                   type = "upper",
                   method = "square", 
                     addCoef.col = "black", 
                     diag = FALSE, 
                     number.cex = .6,
                     tl.col = "black",
                     tl.cex = .9,
                     tl.srt = 45)
```

There appears to several highly correlated potential predictors in our dataset, with strong positive and negative relationships present. There is a strong possibility of multicollinearity in our final models. 

Most starkly, race remains highly correlated with health and SES-related outcomes at a neighborhood level. For example, there is a strong negative correlation between percent population that identifies as white and levels of poverty, unemployment, and late or no prenatal care. This relationship becomes positive when considering percent population that identifies as Hispanic. Surprisingly, proportion of population that identifies as black have weak to no relationship with SES-related outcomes like low educational attainment or poverty, but the strongest relationship with late or no prenatal care (R = 0.54). 

We'll keep these relationships in mind as we interpret our results. Select an outcome from the dropdown and read on!



